Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-12-20 paper.tex
added a few of Barry's edits

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not entirely obvious how pilot symbols should be included. Techniques are available but they are somewhat adhoc~\cite{Cowley_ref_sym_carr_1998}.  This problem does not occur with the least square estimator.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi^2$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi^2$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex conjugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

We only require to consider $\theta$ in the interval $[0, 2\pi)$.  Consider how $\hat{d}_i(\theta)$ changes as $\theta$ varies from $0$ to $2\pi$.  Let $b_i = \hat{d}_i(0)$ and let 
\[
z_i = \angle{y_i} - \hat{u}_i(0) = \angle{y_i} - \round{\angle{y_i}}
\]
be phase difference between the received symbol $y_i$ and the hard decision resulting when $\theta = 0$, i.e. $\round{\angle{y_i}}$.  Then,
\begin{equation}\label{eq:uicombos}
\hat{d}_i(\theta) = 
\begin{cases}
b_i, &  0 \leq \theta < z_i + \frac{\pi}{M} \\
b_i e^{-j2\pi/M}, & z_i + \frac{\pi}{M} \leq \theta < z_i + \frac{3\pi}{M} \\ 
\vdots & \\
b_i e^{-j2\pi k /M}, & z_i + \frac{\pi(2k - 1)}{M} \leq \theta < z_i + \frac{\pi(2k + 1)}{M}  \\ 
\vdots & \\
b_i e^{-j2\pi} = b_i, &  z_i + \frac{\pi(2M - 1)}{M} \leq \theta < 2\pi. \\
\end{cases}
\end{equation}

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $\{y_i, i \in P \cup D\}$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$. %where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  
Put $L = \abs{P \cup D}$ and let $p$ and $d$ be the limits
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i} ,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x) \qquad \text{where,}
\]
\[
h_1(x) = \expect R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\pi,\pi)$, then:
\begin{enumerate}
\item $\sfracpart{\hat{\theta} - \theta_0}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

The proof of Theorem~\ref{thm:consistency} is in Section~\ref{sec:proof-almost-sure} and the proof of Theorem~\ref{thm:normality} is in Section~\ref{sec:proof-asympt-norm}.  Before giving the proofs we discuss the assumptions made by these theorems.  The assumption that $w_1, \dots, w_L$ are circularly symmetric can be relaxed, but this comes at the expense of making the theorem statements more complicated.  If $w_i$ is not circularly symmetric then the distribution of $R_i$ and $\Phi_i$ may depend on $a_0$ and also on the transmitted symbols $\{s_i,i \in P \cup D\}$.  In result, the asymptotic variance described in Theorem~\ref{thm:normality} depends on $a_0$ and $\{s_i,i \in P \cup D\}$, rather than just $\rho_0$.  The circularly symmetric assumption may not always hold in practice, but we feel it provides a sensible trade off between simplicity and generality.

The assumption that $\expect \sabs{w_1}^2 = \expect \sabs{w_i}^2$ is finite implies that $R_i$ has finite variance since $\expect R_i^2 = 1 + \expect \abs{w_i}^2$.  This is required in Theorem~\ref{thm:normality} so that the constants $A_1$, $A_2$, $B_1$ and $B_2$ exist.  We will also use that $R_i$ has finite variance to simplify the proof of Theorem~\ref{thm:consistency} by use of Kolmogorov's strong law of large numbers~\cite[Theorem 2.3.10]{SenSinger_large_sample_stats_1993}

The theorems place conditions on $\sfracpart{\hat{\theta} - \theta_0}_\pi$ rather than directly on $\theta_0  - \theta_0$.  This makes sense because the phases $\theta_0$ and $\theta_0 + 2\pi k$ are equivalent for any integer $k$. So, for example, we expect the phases $0.99\pi$ and $-0.99\pi$ to be close together, the difference between them being $\vert\sfracpart{-0.99\pi - 0.99\pi}_\pi\vert = 0.02\pi$, and not $\vert -0.99\pi - 0.99\pi\vert = 1.98\pi$.

@@

Theorem~\ref{thm:consistency} defines real numbers $p$ and $d$ to represent the proportion of pilot symbols and data symbols in the limit as $L$ goes to infinity.  For Theorem~\ref{thm:normality} we need the slightly stronger condition that 
\[
\frac{\abs{P}}{L} = p + o(L^{-1/2}) \qquad \text{and} \qquad \frac{\abs{D}}{L} = d + o(L^{-1/2})
\] 
where $o(L^{-1/2})$ denotes a number such that $\sqrt{L} o(L^{-1/2})$ goes to zero as $L\rightarrow \infty$.  This stronger condition is required to prove the asymptotic normality of $\sqrt{L}\hat{m}_L$.

The next two sections give proofs for Theorems~\ref{thm:consistency}~and~\ref{thm:normality}.  The proofs make use of various lemmas.  The lemmas are proved in the appendix.

>>>>

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
This problem does not occur with the least square estimator.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi^2$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \big( \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^* \big), \label{eq:SS}
\end{align}
where $*$ denotes complex conjugate.  The minimiser of $SS$ with respect to $a$ as a function of $\{d_i, i \in D\}$ is
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational brevity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ does not depend on the $d_i$.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are at most $M\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

We only require to consider $\theta$ in the interval $[0, 2\pi)$.  Consider how $\hat{d}_i(\theta)$ changes as $\theta$ varies from $0$ to $2\pi$.  Let $b_i = \hat{d}_i(0)$ and let 
\[
z_i = \angle{y_i} - \hat{u}_i(0) = \angle{y_i} - \round{\angle{y_i}}
\]
be the phase difference between the received symbol $y_i$ and the hard decision resulting when $\theta = 0$, i.e. $\round{\angle{y_i}}$.  Then,
\begin{equation}\label{eq:uicombos}
\hat{d}_i(\theta) = 
\begin{cases}
b_i, &  0 \leq \theta < z_i + \frac{\pi}{M} \\
b_i e^{-j2\pi/M}, & z_i + \frac{\pi}{M} \leq \theta < z_i + \frac{3\pi}{M} \\ 
\vdots & \\
b_i e^{-j2\pi k /M}, & z_i + \frac{\pi(2k - 1)}{M} \leq \theta < z_i + \frac{\pi(2k + 1)}{M}  \\ 
\vdots & \\
b_i e^{-j2\pi} = b_i, &  z_i + \frac{\pi(2M - 1)}{M} \leq \theta < 2\pi. \\
\end{cases}
\end{equation}

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = e^{j2\pi/M} - 1$.  In general,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a $O(1)$ arithmetic operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in $O(1)$ operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of $a_0$ is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindices}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

@@

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, circularly symmetric complex random variables with $\expect \abs{w_1}^2$ finite, and let $\{y_i, i \in P \cup D\}$ be given by~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$. %where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  
Put $L = \abs{P \cup D}$ and let $\abs{P}$ and $\abs{D}$ increase in such a way that
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [0,2\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{w_i}{a_0 s_i} ,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x) \qquad \text{where}
\]
\[
h_1(x) = \expect R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \expect R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $G(x)$ is uniquely maximised at $x = 0$ over the interval $[-\pi,\pi)$,
\begin{enumerate}
\item $\sfracpart{\hat{\theta} - \theta_0}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

The proof of Theorem~\ref{thm:consistency} is in Section~\ref{sec:proof-almost-sure} and the proof of Theorem~\ref{thm:normality} is in Section~\ref{sec:proof-asympt-norm}.  Before giving the proofs we discuss the assumptions made by these theorems.  The assumption that $w_1, \dots, w_L$ are circularly symmetric can be relaxed, but this comes at the expense of making the theorem statements more complicated.  If $w_i$ is not circularly symmetric then the distribution of $R_i$ and $\Phi_i$ may depend on $a_0$ and also on the transmitted symbols $\{s_i,i \in P \cup D\}$.  As a result the asymptotic variance described in Theorem~\ref{thm:normality} depends on $a_0$ and $\{s_i,i \in P \cup D\}$, rather than just $\rho_0$.  The circularly symmetric assumption may not always hold in practice, but we feel it provides a sensible trade off between simplicity and generality.

The assumption that $\expect \sabs{w_1}^2 = \expect \sabs{w_i}^2$ is finite implies that $R_i$ has finite variance since $\expect R_i^2 = 1 + \expect \abs{w_i}^2$.  This is required in Theorem~\ref{thm:normality} so that the constants $A_1$, $A_2$, $B_1$ and $B_2$ exist.  We will also use the fact that $R_i$ has finite variance to simplify the proof of Theorem~\ref{thm:consistency} by use of Kolmogorov's strong law of large numbers~\cite{Billingsley1979_probability_and_measure}.

The theorems place conditions on $\sfracpart{\hat{\theta} - \theta_0}_\pi$ rather than directly on $\hat{\theta}  - \theta_0$.  This makes sense because the phases $\theta_0$ and $\theta_0 + 2\pi k$ are equivalent for any integer $k$. So, for example, we expect the phases $0.99\pi$ and $-0.99\pi$ to be close together, the difference between them being $\vert\sfracpart{-0.99\pi - 0.99\pi}_\pi\vert = 0.02\pi$, and not $\vert -0.99\pi - 0.99\pi\vert = 1.98\pi$.

@@

Theorem~\ref{thm:consistency} defines real numbers $p$ and $d$ to represent the proportion of pilot symbols and data symbols in the limit as $L$ goes to infinity.  For Theorem~\ref{thm:normality} we need the slightly stronger condition that 
\[
\frac{\abs{P}}{L} = p + o(L^{-1/2}) \qquad \text{and} \qquad \frac{\abs{D}}{L} = d + o(L^{-1/2}).
\] 
%where $o(L^{-1/2})$ denotes a number such that $\sqrt{L} o(L^{-1/2})$ goes to zero as $L\rightarrow \infty$.  
This stronger condition is required to prove the asymptotic normality of $\sqrt{L}\hat{m}_L$.

The next two sections give proofs of Theorems~\ref{thm:consistency}~and~\ref{thm:normality}.  The proofs make use of various lemmas, which are proved in the appendix.
